Determinism, Complexity, and Predictability in Computer Performance

### 2014-07-16 results.tex
bunch of formatting issues
fixing autoarima command and formatting bugs

The experiments below involve three different prediction methods
applied to time-series data from eight different systems: {\tt
  col\_major}, {\tt 403.gcc}, and the six different regimes of the
{\tt dgesdd} signal in Figure~\ref{fig:svd-ts-colored}.  The objective
of these experiments was to explore how prediction accuracy is related
to WPE.
% 
% , and how that relationship depends on the generating process
% and the prediction method.  
% 
Working from the first 90\% of each signal, we generated a prediction
of the last 10\% using the \naive, AutoARIMA, and LMA prediction methods,
as described in Section~\ref{sec:accuracy}, then calculated the MASE
value of those predictions.  We also calculated the WPE of each time
series using a wordlength chosen via the procedure described in
Section~\ref{sec:meaComplex}.  In order to assess the run-to-run
variability of these results, we repeated all of these calculations on
15 separate trials: i.e., 15 different runs of each program.

@@

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for the full
set of experiments.
\begin{figure}
  \centering
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
  \includegraphics[width=\columnwidth]{figs/new_predictions_vs_entropy3}
\caption{WPE vs. MASE for all trials, methods, and systems---with the
  exception of \svdone, \svdthree, and \svdfive are omitted from the
  top plot for scale reasons.
%
% ; their MASE scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
%   $20.87 \pm 0.192$ respectively.
%
Numerical values, including means and standard deviations of the
errors, can be found in Table~\ref{tab:error}.  The dashed curve [[and
    shaded regions?  need to check this throughout the text and in all
    captions after we finalize the plot]] is the same as in the
previous figure.  }
    \label{fig:wpe_vs_mase_all}
\end{figure}
There are 15 points in each cluster, one for each trial.  (The points
in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
for the corresponding trace in any of the three plots in
Figure~\ref{fig:wpe_vs_mase_all}.)  The WPE values do not vary very
much across trials.  For most traces, the variance in MASE scores is
low as well, resulting in small, tight clusters.  In some
cases---AutoARIMA predictions of {\tt col\_major}, for instance---the
MASE variance is larger, which spreads out the clusters horizontally.
The distribution of the cluster \emph{positions} is quite different in
the three plots.  The mean MASE scores of predictions generated with
the nonlinear LMA method are generally closer to the [[dashed curve]];
the AutoARIMA method clusters are more widely spread, the \naive ~clusters
even more so.  This geometric validation of the second conjecture in
this paper is discussed in more detail in the following paragraphs.

Though \col is a very simple program, its dynamics are actually quite
complicated, as discussed in Section~\ref{sec:intro}.  Recall from the
graphical representations in the left-hand column of
Figure~\ref{fig:forecast-example} that the \naive ~and AutoARIMA
prediction methods do not perform as well on this signal as the
nonlinear LMA method.  The MASE scores of the \naive ~and AutoARIMA
predictions are $0.571 \pm 0.002$ and $0.599 \pm 0.211$, respectively,
across all 15 trials.  That is, these methods are performing only
$\approx 1.7$ times better than the random-walk method, a primitive
strategy that simply uses the current value as the prediction.
However, the WPE value for the \col trials is $0.513 \pm 0.003$, which
is in the center of the complexity spectrum described in
Section~\ref{sec:intro}.

This disparity---WPE values that suggest a high rate of forward
information transfer in the signal, but predictions with comparatively
poor MASE scores---is obvious in the geometry of the top two images in
Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far to
the right of the [[dashed curve]].  \emph{This indicates that these
  methods are not leveraging the available information in the signal}.
The dynamics of \col may be complicated, but they are not
unstructured.  This signal is nonlinear and
deterministic\cite{mytkowicz09}, and if one uses a prediction
technique that is based a nonlinear model (LMA)---rather than a method
that simply predicts the running mean (\naive) or one that uses a
linear model (AutoARIMA)---the MASE score is much better: $0.050 \pm
0.001$.  This prediction is 20 times more accurate than a random-walk
forecast, which is more in line with the level of predictive structure
that the low WPE value suggests is present in the signal.

@@

Note that the \col points (blue {\color{blue}$+$} icons) are clustered
very tightly in the lower left quadrant of the LMA and \naive ~MASE
vs. WPE plots, but spread out horizontally in the AutoARIMA plot.
This is because of the way that these models are built
\cite{autoARIMA}.  If a KPSS test of the time series in question
indicates that it is nonstationary, the AutoARIMA recipe adds an
integration term to the model.  This test gives mixed results in the
case of the \col process, flagging five of the 15 trials as stationary
and ten as nonstationary.  ARIMA models without an integration term
perform more poorly on this signal, which increases the error, thereby
spreading out the points\footnote{We tested this hypothesis by forcing
  the inclusion of an integration term in the five cases where a KPSS
  test indicated that such a term was not needed.  This action removed
  the spread, pushing all 15 of the \col AutoARIMA points in
  Figure~\ref{fig:wpe_vs_mase_all} into a tight cluster.}.
\alert{This paragraph probably needs more, in response to the ref's
  gritch about fitting.  We certainly want to call his attention to
  it, either way.}

The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
  col\_major}.  This indicates that the rate of forward information
transfer of the underlying process is lower, but that time-series data
observed from this system still contain a significant amount of
structure that can, in theory, be used to predict the future course of
the time series.  As mentioned above, though, the effectiveness of any
prediction strategy depends on how well it leverages the information
that is available in the data: methods that use linear models, for
instance, may fail when applied to nonlinear processes.
%
% (This is the basis for the second conjecture above.)
%
This match between model and task is an issue in the \svdfive
experiments as well.  The MASE scores of the \naive ~and AutoARIMA
predictions for this system are $20.870 \pm 0.192$ and $2.370 \pm
0.051$, respectively: that is, 20.87 and 2.37 times worse than a
simple random walk forecast of the same signals\footnote{The \naive
  ~MASE score is large because of the bimodal nature of the
  distribution of the values of the signal, which makes guessing the
  mean a particularly bad strategy.  The same thing is true of the
  \svdthree signal.}.  As before, the positions of these points on a
WPE vs. MASE plot---significantly below and to the right of the shaded
region---should suggest to a practitioner that there is more structure
in this signal than the corresponding method is able to leverage, and
that a different method might do better.  Indeed, for \svdfive, the
LMA method produces a MASE score of $ 0.718\pm 0.048 $ and a cluster
of results that is much closer to the dashed curve on the WPE-MASE
plot.  Again, this validates our second conjecture: the LMA method can
capture and reproduce the way in which the \svdfive system processes
information, but the \naive ~and AutoARIMA prediction methods cannot.

The WPE of \gcc is higher still: $0.943 \pm 0.001$.  This system
transmits very little information forward in time and provides almost
no structure for prediction methods to work with.  Here, the \naive
~method performs best ($MASE=0.951 \pm 0.001$), simply because its
calculation of the mean filters out the noise in the signal.  Since
fitting a hyperplane using least squares should have similar effects,
the fact that LMA outperforms AutoARIMA ($1.530 \pm 0.021$ vs. $1.837
\pm 0.016$) is somewhat counterintuitive.  However, the small amount
of predictive structure that is present in this signal is nonlinear
(cf.,~\cite{mytkowicz09}), and LMA is designed to capture and exploit
that kind of structure.

\svdone---the dark blue segment of
Figure~\ref{fig:svd-ts-colored}---behaves very differently than the
other seven systems in this study.  Though its weighted permutation
entropy is very high ($0.957 \pm 0.016$), two of the three prediction
methods do quite well on this signal, yielding MASE scores of $0.827
\pm 0.076$ (LMA) and $0.714 \pm 0.075$ (AutoARIMA).  This pushes the
corresponding clusters of points in the bottom two images in
Figure~\ref{fig:wpe_vs_mase_all} well above the trend followed by the
other seven signals.  The MASE scores of the predictions that were
produced by the \naive ~method for this system, however, are highly
inconsistent.  The majority of the blue diamond-shaped points on the
top plot in Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE
score of 0.6, which is better than LMA and AutoARIMA.  In five of the
15 \svdone trials, however, there were step changes in the signal.
The \naive ~method has a very difficult time with signals like this,
particularly if there are multiple step changes.  This raised the MASE
scores of these trials, pushing the corresponding points to the
right\footnote{This includes the cluster of three points near MASE
  $\approx 2.5$, as well as two points that are beyond the domain of
  the graph, at MASE 11.2 and 14.8.}, and in turn raising both the
mean and variance of this set of trials.

The effects described in the previous paragraph are also exacerbated
by the way MASE is calculated.  Recall that MASE scores are scaled
\emph{relative to a random-walk forecast}, which uses the current
value as the prediction.  This strategy works very badly on signals
with frequent, large, rapid transitions.  Consider a signal that
oscillates from one end of its range to the other at every step.  A
signal like this will have a low WPE, much like {\tt col\_major}.
However, a random-walk forecast of this signal will be 180 degrees out
of phase with the true continuation.  Since random-walk error appears
in the denominator of the MASE score, this effect can shift points
leftwards on a WPE vs. MASE plot, and that is exactly why the \svdone
clusters for AutoARIMA and LMA are above the dashed curve in
Figure~\ref{fig:wpe_vs_mase_all}.  This time series, which is shown in
closeup in Figure~\ref{fig:svdone-ts},
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\columnwidth]{figs/svdonets2}
\caption{A small portion of the \svdone time series}\label{fig:svdone-ts}
\end{figure}
is not quite to the level of the worst-case signal described above,
but it still poses a serious challenge to random-walk prediction.  It
is dominated by a noisy regime (between $\approx$1.86 and
$\approx$1.88 on the vertical scale in Figure~\ref{fig:svdone-ts}),
punctuated by short excursions above 1.9.  In the former regime, which
makes up more than 80\% of the signal, there are frequent dips to 1.82
and occasional larger dips below 1.8.  These single-point dips are the
bane of random-walk forecasting.  In this particular case, roughly
40\% of the forecasted points are off by the width of the associated
dip, which skews the associated MASE scores.  Signals like this are
also problematic for the \naive ~prediction strategy, since the
outliers have significant influence on the mean.  This compounds the
effect of the skew in the scaling factor and exacerbates the spread in
the \svdone MASE values.  For all of these reasons, we view \svdone as
an outlier and exclude it from the fit calculation of the dashed curve
in Figure~\ref{fig:wpe_vs_mase_best}.

[[Liz to wordsmith: Even though MASE has weaknesses, we concur with
    \cite{mase} that it is an effective way to compare prediction
    error across time series of different lengths and different
    scales.  Point to that paper for extensive evaluation and
    comparison wrt other metrics. We are in the process of repeating
    our own study with other metrics, other forecast methods, and
    time-series data from other systems.  Segue to next section.]]

>>>>

The experiments below involve three different prediction methods
applied to time-series data from eight different systems: {\tt
  col\_major}, {\tt 403.gcc}, and the six different regimes of the
{\tt dgesdd} signal in Figure~\ref{fig:svd-ts-colored}.  The objective
of these experiments was to explore how prediction accuracy is related
to WPE.
% 
% , and how that relationship depends on the generating process
% and the prediction method.  
% 
Working from the first 90\% of each signal, we generated a prediction
of the last 10\% using the \naive, \arima, and LMA prediction methods,
as described in Section~\ref{sec:accuracy}, then calculated the MASE
value of those predictions.  We also calculated the WPE of each time
series using a wordlength chosen via the procedure described in
Section~\ref{sec:meaComplex}.  In order to assess the run-to-run
variability of these results, we repeated all of these calculations on
15 separate trials: i.e., 15 different runs of each program.

@@

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for the full
set of experiments.
\begin{figure}
  \centering
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
  \includegraphics[width=\columnwidth]{figs/new_predictions_vs_entropy3}
\caption{WPE vs. MASE for all trials, methods, and systems---with the
  exception of \svdone, \svdthree, and \svdfive are omitted from the
  top plot for scale reasons.
%
% ; their MASE scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
%   $20.87 \pm 0.192$ respectively.
%
Numerical values, including means and standard deviations of the
errors, can be found in Table~\ref{tab:error}.  The dashed curve [[and
    shaded regions?  need to check this throughout the text and in all
    captions after we finalize the plot]] is the same as in the
previous figure.  }
    \label{fig:wpe_vs_mase_all}
\end{figure}
There are 15 points in each cluster, one for each trial.  (The points
in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
for the corresponding trace in any of the three plots in
Figure~\ref{fig:wpe_vs_mase_all}.)  The WPE values do not vary very
much across trials.  For most traces, the variance in MASE scores is
low as well, resulting in small, tight clusters.  In some
cases---\arima predictions of {\tt col\_major}, for instance---the
MASE variance is larger, which spreads out the clusters horizontally.
The distribution of the cluster \emph{positions} is quite different in
the three plots.  The mean MASE scores of predictions generated with
the nonlinear LMA method are generally closer to the [[dashed curve]];
the \arima method clusters are more widely spread, the \naive ~clusters
even more so.  This geometric validation of the second conjecture in
this paper is discussed in more detail in the following paragraphs.

Though \col is a very simple program, its dynamics are actually quite
complicated, as discussed in Section~\ref{sec:intro}.  Recall from the
graphical representations in the left-hand column of
Figure~\ref{fig:forecast-example} that the \naive ~and \arima
prediction methods do not perform as well on this signal as the
nonlinear LMA method.  The MASE scores of the \naive ~and \arima
predictions are $0.571 \pm 0.002$ and $0.599 \pm 0.211$, respectively,
across all 15 trials.  That is, these methods are performing only
$\approx 1.7$ times better than the random-walk method, a primitive
strategy that simply uses the current value as the prediction.
However, the WPE value for the \col trials is $0.513 \pm 0.003$, which
is in the center of the complexity spectrum described in
Section~\ref{sec:intro}.

This disparity---WPE values that suggest a high rate of forward
information transfer in the signal, but predictions with comparatively
poor MASE scores---is obvious in the geometry of the top two images in
Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far to
the right of the [[dashed curve]].  \emph{This indicates that these
  methods are not leveraging the available information in the signal}.
The dynamics of \col may be complicated, but they are not
unstructured.  This signal is nonlinear and
deterministic\cite{mytkowicz09}, and if one uses a prediction
technique that is based a nonlinear model (LMA)---rather than a method
that simply predicts the running mean (\naive) or one that uses a
linear model (\arima)---the MASE score is much better: $0.050 \pm
0.001$.  This prediction is 20 times more accurate than a random-walk
forecast, which is more in line with the level of predictive structure
that the low WPE value suggests is present in the signal.

@@

Note that the \col points (blue {\color{blue}$+$} icons) are clustered
very tightly in the lower left quadrant of the LMA and \naive ~MASE
vs. WPE plots, but spread out horizontally in the \arima plot.
This is because of the way that these models are built
\cite{autoARIMA}.  If a KPSS test of the time series in question
indicates that it is nonstationary, the \arima recipe adds an
integration term to the model.  This test gives mixed results in the
case of the \col process, flagging five of the 15 trials as stationary
and ten as nonstationary.  ARIMA models without an integration term
perform more poorly on this signal, which increases the error, thereby
spreading out the points\footnote{We tested this hypothesis by forcing
  the inclusion of an integration term in the five cases where a KPSS
  test indicated that such a term was not needed.  This action removed
  the spread, pushing all 15 of the \col \arima points in
  Figure~\ref{fig:wpe_vs_mase_all} into a tight cluster.}.
\alert{This paragraph probably needs more, in response to the ref's
  gritch about fitting.  We certainly want to call his attention to
  it, either way.}

The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
  col\_major}.  This indicates that the rate of forward information
transfer of the underlying process is lower, but that time-series data
observed from this system still contain a significant amount of
structure that can, in theory, be used to predict the future course of
the time series.  As mentioned above, though, the effectiveness of any
prediction strategy depends on how well it leverages the information
that is available in the data: methods that use linear models, for
instance, may fail when applied to nonlinear processes.
%
% (This is the basis for the second conjecture above.)
%
This match between model and task is an issue in the \svdfive
experiments as well.  The MASE scores of the \naive ~and \arima
predictions for this system are $20.870 \pm 0.192$ and $2.370 \pm
0.051$, respectively: that is, 20.87 and 2.37 times worse than a
simple random walk forecast of the same signals\footnote{The \naive
  ~MASE score is large because of the bimodal nature of the
  distribution of the values of the signal, which makes guessing the
  mean a particularly bad strategy.  The same thing is true of the
  \svdthree signal.}.  As before, the positions of these points on a
WPE vs. MASE plot---significantly below and to the right of the shaded
region---should suggest to a practitioner that there is more structure
in this signal than the corresponding method is able to leverage, and
that a different method might do better.  Indeed, for \svdfive, the
LMA method produces a MASE score of $ 0.718\pm 0.048 $ and a cluster
of results that is much closer to the dashed curve on the WPE-MASE
plot.  Again, this validates our second conjecture: the LMA method can
capture and reproduce the way in which the \svdfive system processes
information, but the \naive ~and \arima prediction methods cannot.

The WPE of \gcc is higher still: $0.943 \pm 0.001$.  This system
transmits very little information forward in time and provides almost
no structure for prediction methods to work with.  Here, the \naive
~method performs best ($MASE=0.951 \pm 0.001$), simply because its
calculation of the mean filters out the noise in the signal.  Since
fitting a hyperplane using least squares should have similar effects,
the fact that LMA outperforms \arima ($1.530 \pm 0.021$ vs. $1.837
\pm 0.016$) is somewhat counterintuitive.  However, the small amount
of predictive structure that is present in this signal is nonlinear
(cf.,~\cite{mytkowicz09}), and LMA is designed to capture and exploit
that kind of structure.

\svdone---the dark blue segment of
Figure~\ref{fig:svd-ts-colored}---behaves very differently than the
other seven systems in this study.  Though its weighted permutation
entropy is very high ($0.957 \pm 0.016$), two of the three prediction
methods do quite well on this signal, yielding MASE scores of $0.827
\pm 0.076$ (LMA) and $0.714 \pm 0.075$ (\arima).  This pushes the
corresponding clusters of points in the bottom two images in
Figure~\ref{fig:wpe_vs_mase_all} well above the trend followed by the
other seven signals.  The MASE scores of the predictions that were
produced by the \naive ~method for this system, however, are highly
inconsistent.  The majority of the blue diamond-shaped points on the
top plot in Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE
score of 0.6, which is better than LMA and \arima.  In five of the
15 \svdone trials, however, there were step changes in the signal.
The \naive ~method has a very difficult time with signals like this,
particularly if there are multiple step changes.  This raised the MASE
scores of these trials, pushing the corresponding points to the
right\footnote{This includes the cluster of three points near MASE
  $\approx 2.5$, as well as two points that are beyond the domain of
  the graph, at MASE 11.2 and 14.8.}, and in turn raising both the
mean and variance of this set of trials.

The effects described in the previous paragraph are also exacerbated
by the way MASE is calculated.  Recall that MASE scores are scaled
\emph{relative to a random-walk forecast}, which uses the current
value as the prediction.  This strategy works very badly on signals
with frequent, large, rapid transitions.  Consider a signal that
oscillates from one end of its range to the other at every step.  A
signal like this will have a low WPE, much like {\tt col\_major}.
However, a random-walk forecast of this signal will be 180 degrees out
of phase with the true continuation.  Since random-walk error appears
in the denominator of the MASE score, this effect can shift points
leftwards on a WPE vs. MASE plot, and that is exactly why the \svdone
clusters for \arima and LMA are above the dashed curve in
Figure~\ref{fig:wpe_vs_mase_all}.  This time series, which is shown in
closeup in Figure~\ref{fig:svdone-ts},
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\columnwidth]{figs/svdonets2}
\caption{A small portion of the \svdone time series}\label{fig:svdone-ts}
\end{figure}
is not quite to the level of the worst-case signal described above,
but it still poses a serious challenge to random-walk prediction.  It
is dominated by a noisy regime (between $\approx$1.86 and
$\approx$1.88 on the vertical scale in Figure~\ref{fig:svdone-ts}),
punctuated by short excursions above 1.9.  In the former regime, which
makes up more than 80\% of the signal, there are frequent dips to 1.82
and occasional larger dips below 1.8.  These single-point dips are the
bane of random-walk forecasting.  In this particular case, roughly
40\% of the forecasted points are off by the width of the associated
dip, which skews the associated MASE scores.  Signals like this are
also problematic for the \naive ~prediction strategy, since the
outliers have significant influence on the mean.  This compounds the
effect of the skew in the scaling factor and exacerbates the spread in
the \svdone MASE values.  For all of these reasons, we view \svdone as
an outlier and exclude it from the fit calculation of the dashed curve
in Figure~\ref{fig:wpe_vs_mase_best}.

[[Liz to wordsmith: Even though MASE has weaknesses, we concur with
    \cite{MASE} that it is an effective way to compare prediction
    error across time series of different lengths and different
    scales.  Point to that paper for extensive evaluation and
    comparison wrt other metrics. We are in the process of repeating
    our own study with other metrics, other forecast methods, and
    time-series data from other systems.  Segue to next section.]]
